\stktokens{}: Enforcing Well-Bracketed Control Flow and Stack Encapsulation Using Linear Capabilities

### 2020-12-19 popl-2019-journal-v/proving-full-abstraction.tex
Start polishing and shortening section 5.

\begin{jversion}
Assumptions about the system that relatedness is predicated on are gathered in (Kripke) worlds.
To a first approximation, a world is a semantic model of the memory.
In its simplest form, it is a collection of invariants that the memory must satisfy.
The invariants of a world can vary in complexity and expressiveness depending on the application.
The possible contents of the memory also influence what the world looks like.
For instance, the uniqueness of the linear capabilities on \trgcm{} and \srccm{} is modelled by the worlds using a form of memory ownership assumptions.
In order to relate \trgcm{} and \srccm{}, we model all the features of \srccm{} in the world which means we have to model:
\begin{itemize}
\item Three kinds of memory: heap, stack of local frames, and free stack
\item Linearity
\item Call stack
\item Seals
\end{itemize}
The three kinds of memory are modelled by having three sub-worlds where each sub-world is its own little world in a traditional sense.
Linearity is modelled by adding ownership to certain parts of the world that capabilities can take ownership of ensuring that they are the sole reference for that part of memory.
Memory satisfaction (the relation that decides whether two memories are related in the world) models the call stack by ensuring that the memory is actually shaped like a stack.
Finally, the seals are modelled by seal invariants that make sure that seals are only used on permitted sealables.
In the following, we present the world and go into details about how each of the four features are modelled.

@@

The sub-worlds are partial maps from names $\RegionName$ (not to be confused with register names), modelled as natural numbers, to regions, which model a form of invariant.
A region describes a collection of related memory segments, so it is simply represented as a relation over memory segments (i.e.\ a relation in $\Rel{\MemSeg \times \MemSeg}$).
Intuitively, we want to be able to say that two memory segments are related when their content is related.
That is, for every address in the two memory segments, the words that reside there must be related.
In Section~\ref{subsec:logical-relation}, we define precisely what it means for words to be related, but for now we provide some intuition.
If we have two integers, then they are related when they are equal.
If we instead have two capabilities, then they should also be related if they, in some sense, are equal.
But what does it mean for capabilities to be \textit{equal}?
Intuitively, it should mean that the capabilities give you the same authority, for instance, if you have two executable capabilities, they should observably perform the same computation.
As a capability points to a piece of memory, the authority of the capability depends on the contents of that memory.
In other words to say whether two capabilities are related, we must know the possible contents of the memory.
The world expresses the possible memory contents, so our regions are world indexed, i.e.
\[
  \Worldh = \RegionName \parfun (\World \fun \Rel{\MemSeg \times \MemSeg})
\]
At this point, we can see that we have constructed a recursive domain equation.
If we inline $\Worldh$ in $\World$, then we have a circular equation with no solution because the self-reference happens in a negative position.
Luckily, we can solve circular equations if we move to a different domain.
For now, we will ignore the problem and return to the issue in Section~\ref{sec:rec-dom-eq}.

For the sake of readability, we introduce the following notation
\begin{gather*}
  \pwheap = \pi_1(W)\\
  \pwpriv = \pi_2(W)\\
  \pwfree = \pi_3(W)
\end{gather*}
\subsubsection{Linearity}
% Linearity
The linear capabilities of \srccm{} and \trgcm{} guarantee that they have the sole authority over the memory they reference\footnote{Under the assumption that the system was initialised with unique linear capabilties.}.
In order to model this uniqueness, we need to keep track of which parts of memory are uniquely referenced and make sure that only one linear capability references the unique parts.
%% Bookkeeping done by worlds
We use the world to keep track of what parts of memory must be uniquely referenced by having two kinds of regions: shared and spatial.
If a memory segment is governed by a shared region, then normal capabilities may reference it.
On the other hand, if a memory segment is governed by a spatial region, then only a linear capability may reference it.

%%% spatial/spatial_owned
However, spatial regions are not enough.
A world which contains a spatial region, represents the assumption of exclusive ownership of the represented memory fragment.
In other words, a linear capability for this memory may be valid in such a world.
However, when we do not possess exclusive ownership of a memory fragment, that does not mean we should be entirely unaware of the memory's existence.
To represent knowledge about existence of a spatial region, without requiring its exclusive ownership, we have $\spatial$ regions.
Specifically, we add tags $\spatialo$ and $\spatial$ to the spatial regions:
\[
  \Regions = \left\{
  \begin{array}{l}
    \{\spatialo \} \times (\World \fun \Rel{\MemSeg \times \MemSeg}) \cup \\ 
    \{\spatial \} \times (\World \fun \Rel{\MemSeg \times \MemSeg})  \\
  \end{array} \right.
\]
For readability, we also add a tag $\pure$ to the shared regions:
\[
  \Regionh = \{\pure \} \times (\World \fun \Rel{\MemSeg \times \MemSeg}) 
\]

A $\spatial$ region is a shadow copy of a spatial region in the sense that it specifies part of memory, but it does not give the right to reference that part of memory.
This does not mean that the memory is necessarily not referenced.
When we discuss compatibility of worlds w.r.t. ownership in Section~\ref{subsubsec:joining-worlds}, we will see that a world with a shadow region can be combined with a world that has the same region as $\spatialo$.
This setup will allow ownership of a memory fragment to reside with linear capabilities which are not themselves stored in the memory fragment but elsewhere in memory or in a register.
In such a setup, the region representing the memory fragment will exist only as $\spatial$ in the world where the memory fragment's contents is valid, but it will exist as $\spatialo$ in the world where the linear capability owning the fragment is valid.

% What parts of memory are linear
We will extend the regions further in Sections~\ref{subsubsec:seals}~and~\ref{subsubsec:ft-and-revocation}, but for now we continue the definitions of the three sub-worlds.
The sub-world $\Worldh$ specifies the heap memory which can be referenced by both linear and normal capabilities, so it should contain both shared and spatial regions.
For this reason, it is defined as
\[
  \Worldh = \RegionName \parfun (\Regionh \cup \Regions)
\]
\srccm{} internalizes the \stktokens{} stack, so it can only be referenced by linear capabilities which means that the two stack regions should only have spatial regions.
For instance the $\Worldfs$ is defined as
\[
  \Worldfs = \RegionName \parfun \Regions
\]
$\Worlds$ can also only be referenced by linear capabilities, so it should also only have spatial regions.
$\Worlds$ not only models the memory contents of the local stack frames, it also models the call stack that consists of the stack frames.
Conceptually, this means that each of the stack frames is connected with a return point in some code.
In a traditional C calling convention, the code return point would even be stored in the stack frame.
However, with \stktokens{} a stack frame does not contain any information about the corresponding code return point.
Instead, the stack frame and code return point are connected by the capabilities that reference them as they are sealed with the same seal and together then constitute the sealed return pair.
In order to ensure that each call actually returns to the correct point of the code, we must still include the address of the return point in our model.
To this end, each shared region in $\Worlds$ must be paired with a return address:
\[
\Worlds = \RegionName \parfun (\Regions \times \Addr)
\]

% Ensuring linearity
The spatial regions add the necessary bookkeeping to the worlds to model linear capabilities.
The logical relation presented in Section~\ref{subsec:logical-relation} uses this bookkeeping to ensure that linear capabilities \emph{uniquely} reference part of memory.

Given a world, we want to be able to express that a capability that is otherwise valid with respect to the world is not linear or indirectly depends on a linear capability.
This is expressed by stripping the world of all its ownership which corresponds to replacing all $\spatialo$ regions with $\spatial$ regions and then requiring that the capability is valid w.r.t.\ this new world.
We refer to this as the shared part of the world and define the function $\purePart{}$ which turns all spatial regions into shadow copies.
\begin{definition}[The shared part of a world]
  \label{def:purePart}
  For any world $W$, we define
  \begin{align*}
    \purePart{W} &\defeq (\purePart{\pwheap},\purePart{\pwpriv},\purePart{\pwfree})\\
    \purePart{W_\var{heap}} &\defeq \lambda r\ldotp
                       \begin{cases}
                         (\spatial,\var{H}) & \text{if } W_\var{heap}(r) = (\spatialo,\var{H})\\
                         W_{\var{heap}}(r) & \text{otherwise}
                       \end{cases}\\
    \purePart{W_\var{call\_stk}} &\defeq \lambda r\ldotp
                       \begin{cases}
                         ((\spatial,\var{H}),\opc) & \text{if } W_\var{call\_stk}(r) = ((\spatialo,\var{H}),\opc)\\
                         W_\var{call\_stk}(r) & \text{otherwise}
                       \end{cases}\\
    \purePart{W_\var{free}} &\defeq \lambda r\ldotp
                       \begin{cases}
                         (\spatial,\var{Hs}) & \text{if } W_{\var{free}}(r) = (\spatialo,\var{Hs})\\
                         W_{\var{free}}(r) & \text{otherwise}
                       \end{cases}
  \end{align*}
\end{definition}
\subsubsection{Seals}
\label{subsubsec:seals}
So far, the world represents a collection of assumptions on the memory contents that value correctness may depend on.
However, value correctness may depend on other assumptions.
Specifically, \stktokens{} has certain assumption on the seals used for return capabilities and closures.
For instance, a return seal must only be used to seal the return pointer of one specific return point.
Therefore, in addition to a relation on memory segments, some regions also carry a \emph{seal interpretation function} that relates the sealables that may be sealed with a given seal.
  \[
    \Seal \parfun \World \fun \Rel{\SealableCaps \times \SealableCaps}
  \]
%% Shared region has a seal indexed invariant that specifies what sealables can be sealed with a specific seal
  In \stktokens{}, once a seal has been used for a specific purpose (e.g.\ for sealing return capability pairs for a specific call site), it can never be reused for a different purpose.
  This is because there may still be copies of return capabilities out there, signed with the seal.
  This situation is similar to the situation for non-linear memory capabilities, so we only allow shared regions to carry seal interpretation functions, as we will see that those regions can never be revoked in future worlds.
\begin{multline*}
  \Regionh = 
  \{\pure \} \times (\World \fun \Rel{\MemSeg\times\MemSeg}) \times \\
  (\Seal \parfun \World \fun \Rel{\SealableCaps \times \SealableCaps})
\end{multline*}
We also refer to the seal interpretation function as the seal invariant, and we will refer to the memory relation as the memory invariant or just invariant when it is unambiguous.

@@

\subsubsection{Future worlds and revocation}
\label{subsubsec:ft-and-revocation}
% Memory evolves over time, we need to model this.
Very often, relatedness of two capabilities does not change if extra assumptions in the system are added.
For example, two related capabilities remain related when an extra invariant is added on unrelated memory, or when a stack frame that it does not reference is dropped.
In Kripke logical relations, this kind of assumption changes, that do not invalidate the relatedness of values, are modelled by the future world relation $\future$.
The future world relation can also be thought of as the model of allowed changes in memory over time.
We will say that $W'$ is a future world of $W$ if $W' \future W$.

Safety of capabilities should be defined, so that it is monotone with respect to the future world relation.
This means that the system assumptions that capabilities rely on for safety are defined such that capabilities remain safe during execution.
The same holds for the memory invariants stored in the world: capabilities stored in memory should remain valid under allowed changes in system assumptions.
This means we require the world-indexed memory invariants to be monotone in the world.
In other words, a pair of memory segments that are related now will stay related in future worlds.
This changes the spatial regions as follows:
\[
  \Regions = \left\{
  \begin{array}{l}
    \{\spatial \} \times (\World \monfun \Rel{\MemSeg \times \MemSeg}) \cup \\
    \{\spatialo \} \times (\World \monfun \Rel{\MemSeg \times \MemSeg}) \cup \\
    \{\revoked \}
  \end{array} \right.
\]
We make a similar change to $\Regionh$. The seal invariants are monotone as well.

Kripke future world relations usually allow extending worlds with extra assumptions, or take steps in protocols that the system was designed to support from the start.
However in our setting, we sometimes allow dropping assumptions, namely when linearity tells us that no value in the system depends on this assumption any more.
Specifically, if we have the only linear capability for a piece of memory, then we can be sure that there are no other capabilities for the same memory which makes it safe to repurpose the memory and drop or replace the previous assumption.
We mark dropped regions as revoked in the world following \citet{ahmed_2004} and \citet{thamsborg_2011} which for all intents and purposes corresponds to actually dropping the region.

We add a $\revoked$ tag to the spatial regions $\Regions$ which results in
\[
  \Regions = \left\{
  \begin{array}{l}
    \{\spatial \} \times (\World \fun \Rel{\MemSeg \times \MemSeg}) \cup \\
    \{\spatialo \} \times (\World \fun \Rel{\MemSeg \times \MemSeg}) \cup \\
    \{\revoked \}
  \end{array} \right.
\]
The $\spatialo$-region signifies that a linear capability \emph{may} depend on it which means that it cannot be revoked.
On the other hand, no capability can depend on a $\spatial$-region, so it can be safely revoked.
This may be confusing as a $\spatial$-region does not mean that you ``have'' the capability, but we just explained that having a linear capability means that you can repurpose the memory it points to.
As we will see later, the logical relation splits worlds with respect to their ownership, so each capability can get its own world with unique ownership to depend on.
A repurposed linear capability gets an entirely new world to depend on.
The new world constructed for the repurposed linear capability would have a revoked region in place of the $\spatialo$-region, but the new world would not be a future world of the world the linear capability depended on before.
After all, a future world has to respect old invariants, but a repurposed capability needs to depend on new invariants.
While the repurposed capability gets a new world to depend on, all other capabilities in the system depend on the same invariants, so they must be valid with respect to worlds that respect the old invariants, i.e. future worlds.
Further, the future worlds must be compatible (agree on everything but ownership) with the new world.
This means that all the $\spatial$-regions must be replaced with $\revoked$ regions which is why $\spatial$ regions can be turned into $\revoked$ regions.

We define the future world relation in terms of a future region relation which is displayed in Figure~\ref{fig:ft-reg-rel}.
Apart from being revoked, a region can stay the same, or a $\spatial$ region can become $\spatialo$ which models the affinity of the linear capabilities.
Specifically, the linear capabilities in \trgcm{} and \srccm{} can be dropped by overwriting them in registers or memory which makes the linear capabilities affine.
A linear capability is safe with respect to a world with some $\spatialo$ regions.
However, this world, and thus the $\spatialo$ regions, is no longer needed if the linear capability is dropped, so other worlds are free to claim the $\spatialo$ region in place of their $\spatial$ region.
% describe why "Spatial -> spatialo" ~ affine or at least why we have it?
\begin{figure}[htb]
  \centering
  \begin{mathpar}
  \inferrule{ r \in \Regions \cup \Regionh }{ r \future r}
  \and
  \inferrule{ }{ \revoked \future (\spatial,\_)}
  \and
  \inferrule{ }{ (\spatialo,H) \future (\spatial,H)}
\end{mathpar}
  \caption{Future region relation.}
  \label{fig:ft-reg-rel}
\end{figure}
The above repurposing discussion is exemplified in \figurename~\ref{fig:world-example}.
The Figure displays two capabilities $c_{\var{normal}}$ and $c_{\var{linear}}$ that are normal and linear, respectively.
The linear capability is valid with respect to $W_1$ which has the necessary $\spatialo$ region, and the normal capability is valid with respect to $W_2$ which has a $\pure$ region.
The two worlds are compatible as $W_2$ has a $\spatial$ region that matches the $\spatialo$ region of $W_1$.
When the linear capability is repurposed, it must be reflected by the worlds $W_1'$ and $W_2'$.
In both new worlds, the $r_2$ region is replaced with a $\revoked$ region.
$W_1'$ has a new $\spatialo$ region at $r_3$, and $W_2'$ gets a matching $\spatial$ region.
The new world for the linear capability, $W_1'$, is not a future world of $W_1$ as it replaces a $\spatialo$ region with a $\revoked$ region which is not allowed by the future region relation.
The new world for the normal capability, $W_2'$, is a future world of $W_2$ as the future region relation allows $\spatial$ regions to be revoked.
The normal capability $c_{\var{normal}}$ remains valid in $W_2'$ as the $\pure$ region it depends on is monotone with respect to the future world relation.
\begin{figure}\centering

@@

\subsubsection{Joining worlds}
\label{subsubsec:joining-worlds}
The world serves multiple purposes as it is both a specification of memory contents as well as a specification of authority.
This is best seen in the operators used to join worlds.
First when we see the world as a memory specification, we have a pretty standard join $\uplus$ that simply requires the worlds to have different region names.\begin{definition}[World disjoint union $\uplus$]
  Given worlds $W_1$, $W_2$, $W$
  \[
    W_1 \uplus W_2 = W
    \text{ iff }
    \begin{array}[t]{l}
      \dom(\pwheap) = \dom(\pwheap[W_1]) \uplus \dom(\pwheap[W_2]) \tand \\
      \dom(\pwfree) = \dom(\pwfree[W_1]) \uplus \dom(\pwfree[W_2]) \tand \\
      \dom(\pwpriv) = \dom(\pwpriv[W_1]) \uplus \dom(\pwpriv[W_2]) \\
    \end{array}
  \]
\end{definition}
The $\uplus$ world join does not guarantee that the result is a sensible world with respect to authority or memory specification.
In \sectionname~\ref{subsubsec:mem-sat}, we define memory satisfaction which also acts as a well-formedness judgement.

When we view the world as a specification of authority, then the world join need to respect the region ownership.
That is, when we join the authority of two worlds, then the ownership of the two worlds should not overlap.
This is expressed by the $\oplus$ operator.
\begin{definition}[$\oplus$, disjoint union of ownership]
  $W_1 \oplus W_2 = W$ iff
  \[
  \begin{array}[t]{l}
    \dom(\pwheap) = \dom(\pwheap[W_1]) = \dom(\pwheap[W_2]) \tand \\
    \dom(\pwfree) = \dom(\pwfree[W_1]) = \dom(\pwfree[W_2]) \tand \\
    \dom(\pwpriv) = \dom(\pwpriv[W_1]) = \dom(\pwpriv[W_2]) \tand \\
    \forall r \in \dom(\pwheap) \ldotp \pwheap(r) = \pwheap[W_1](r) \oplus \pwheap[W_2](r) \tand \\
    \forall r \in \dom(\pwfree) \ldotp \pwfree(r) = \pwfree[W_1](r) \oplus \pwfree[W_2](r) \tand \\
    \forall r \in \dom(\pwpriv) \ldotp \pi_1(\pwpriv(r)) = \pi_1(\pwpriv[W_1](r)) \oplus \pi_1(\pwpriv[W_2](r))
  \end{array}
  \]
  where $\oplus$ for regions is defined as
\begin{align*}
  (\pure,H,H_{\mathrm{seal}}) \oplus (\pure,H,H_{\mathrm{seal}}) =  & \; (\pure,H,H_{\mathrm{seal}}) \\
  (\spatial,H) \oplus (\spatial,H) =  & \; (\spatial,H) \\
  \revoked \oplus \revoked = & \; \revoked \\
  (\spatialo,H) \oplus (\spatial,H) = & \; (\spatial,H) \oplus (\spatialo,H)\\
                                           =  & \; (\spatialo,H)
\end{align*}
\end{definition}
The $\oplus$ operator, like the $\uplus$ operator, does not guarantee that the resulting world is sensible.
For example, the resulting world might have several regions that impose conflicting requirements on a single memory location.
However, when we know that a certain memory satisfies the world (as a memory specification, see Section~\ref{subsubsec:mem-sat}), then we can be sure that the world's specifications are non-contradictory.

>>>>

\begin{jversion}
These assumptions are explictly represented using (Kripke) worlds.
To a first approximation, a world is a semantic model of the memory.
In its simplest form, it is a collection of invariants that the memory must satisfy.
The invariants of a world can vary in complexity and expressiveness depending on the application.
% The possible contents of the memory also influence what the world looks like.
% For instance, the uniqueness of the linear capabilities on \trgcm{} and \srccm{} is modelled by the worlds using a form of memory ownership assumptions.
In order to relate \trgcm{} and \srccm{}, we model all the features of \srccm{} in the world which means we have to model:
\begin{itemize}
\item Three kinds of memory: heap, stack of local frames, and free stack
\item Linearity
\item Call stack
\item Seals
\end{itemize}
The three kinds of memory are modelled by having three sub-worlds where each sub-world is its own little world in a traditional sense.
Linearity is modelled by adding ownership to certain parts of the world which can be claimed
exclusively by capabilities.
Memory satisfaction (the relation that decides whether two memories are related in the world) models the call stack by ensuring that the memory is actually shaped like a stack.
Finally, the seals are modelled by seal invariants that make sure that seals are only used on permitted sealables.
In the following, we present the world and go into details about how each of the four features are modelled.

@@

The sub-worlds are partial maps from names $\RegionName$ (not to be confused with register names), modelled as natural numbers, to regions, which model a form of invariant.
Intuitvely, a region is simply a relation over memory segments (i.e.\ a relation in $\Rel{\MemSeg \times \MemSeg}$).
However, this is not sufficient, because the validity of memory contents must often depend on the world itself.
In other words, our regions must be world indexed, i.e.
\[
  \Worldh = \RegionName \parfun (\World \fun \Rel{\MemSeg \times \MemSeg})
\]
At this point, we can see that we have constructed a recursive domain equation.
If we inline $\Worldh$ in $\World$, then we have a circular equation with no solution because the self-reference happens in a negative position.
However, in Section~\ref{sec:rec-dom-eq}, we will explain how we can solve the circular equation by moving to a different domain.

For the sake of readability, we introduce the following notation
\begin{gather*}
  \pwheap = \pi_1(W)\\
  \pwpriv = \pi_2(W)\\
  \pwfree = \pi_3(W)
\end{gather*}
\subsubsection{Linearity}
% Linearity
The linear capabilities of \srccm{} and \trgcm{} guarantee sole authority over the memory they reference.
To model this uniqueness, we need to keep track of which parts of memory are uniquely referenced and make sure that only one linear capability references them.
%% Bookkeeping done by worlds
We use the world to do this by having two kinds of regions: shared and spatial.
Memory governed by shared and spatial regions can only be references by non-linear and linear capabilities respectively.

%%% spatial/spatial_owned
However, spatial regions are not enough.
A world which contains a spatial region, represents the assumption of exclusive ownership of the corresponding memory.
In other words, a linear capability for this memory will be valid only in such a world.
However, that does not mean all other worlds should be entirely unaware of the memory's existence.
To represent knowledge about existence of a spatial region, without requiring its exclusive ownership, we have $\spatial$ regions.
Specifically, we add tags $\spatialo$ and $\spatial$ to the spatial regions:
\[
  \Regions = \left\{
  \begin{array}{l}
    \{\spatialo \} \times (\World \fun \Rel{\MemSeg \times \MemSeg}) \cup \\ 
    \{\spatial \} \times (\World \fun \Rel{\MemSeg \times \MemSeg})  \\
  \end{array} \right.
\]
For readability, we also add a tag $\pure$ to the shared regions:
\[
  \Regionh = \{\pure \} \times (\World \fun \Rel{\MemSeg \times \MemSeg}) 
\]

A $\spatial$ region is a shadow copy of a spatial region in the sense that it specifies part of memory, but it does not give the right to reference that part of memory.
We will see in Section~\ref{subsubsec:joining-worlds} that a world with a shadow region is compatible with a world that has the same region as $\spatialo$.
This setup will allow ownership of a memory fragment to reside with linear capabilities which are not themselves stored in the memory fragment but elsewhere in memory or in a register.
In such a setup, the region representing the memory fragment will exist only as $\spatial$ in the world where the memory fragment's contents is valid, but it will exist as $\spatialo$ in the world where the linear capability owning the fragment is valid.

% What parts of memory are linear
We will extend the regions further in Sections~\ref{subsubsec:seals}~and~\ref{subsubsec:ft-and-revocation}, but for now we continue the definitions of the three sub-worlds.
The sub-world $\Worldh$ specifies the heap memory which can be referenced by both linear and normal capabilities, so it should contain both shared and spatial regions.
For this reason, it is defined as
\[
  \Worldh = \RegionName \parfun (\Regionh \cup \Regions)
\]
\srccm{} internalizes the \stktokens{} stack, which will only be referenced by linear capabilities, so the two stack regions only need spatial and shadow regions.
For instance the $\Worldfs$ is defined as
\[
  \Worldfs = \RegionName \parfun \Regions
\]
$\Worlds$ not only models the memory contents of the local stack frames, it also models the call stack itself and specifically the code location for the return point of every frame.
In a traditional C calling convention, this return location would simply be stored in the stack frame, but \stktokens{} does not require this.
It is the code part of the return capability pair that references this return location and it is connected to the stack frame by the fact that it is sealed with the same seal as the linear capability that owns the frame: the data part of the return capability pair.
To ensure and remember that all of this is setup correctly, the world will also keep track of the return location for every frame:
\[
\Worlds = \RegionName \parfun (\Regions \times \Addr)
\]
It will in fact also remember the seal used, as we will see in the next Section.

Given a world, we want to be able to express that a capability that is otherwise valid with respect to the world is not linear or indirectly depends on a linear capability.
This is expressed by stripping the world of all its ownership which corresponds to replacing all $\spatialo$ regions with $\spatial$ regions and then requiring that the capability is valid w.r.t.\ this new world.
We refer to this as the shared part of the world and define the function $\purePart{}$ which turns all spatial regions into shadow copies.
\begin{definition}[The shared part of a world]
  \label{def:purePart}
  For any world $W$, we define
  \begin{align*}
    \purePart{W} &\defeq (\purePart{\pwheap},\purePart{\pwpriv},\purePart{\pwfree})\\
    \purePart{W_\var{heap}} &\defeq \lambda r\ldotp
                       \begin{cases}
                         (\spatial,\var{H}) & \text{if } W_\var{heap}(r) = (\spatialo,\var{H})\\
                         W_{\var{heap}}(r) & \text{otherwise}
                       \end{cases}\\
    \purePart{W_\var{call\_stk}} &\defeq \lambda r\ldotp
                       \begin{cases}
                         ((\spatial,\var{H}),\opc) & \text{if } W_\var{call\_stk}(r) = ((\spatialo,\var{H}),\opc)\\
                         W_\var{call\_stk}(r) & \text{otherwise}
                       \end{cases}\\
    \purePart{W_\var{free}} &\defeq \lambda r\ldotp
                       \begin{cases}
                         (\spatial,\var{Hs}) & \text{if } W_{\var{free}}(r) = (\spatialo,\var{Hs})\\
                         W_{\var{free}}(r) & \text{otherwise}
                       \end{cases}
  \end{align*}
\end{definition}

\subsubsection{Seals}
\label{subsubsec:seals}
So far, the world represents a collection of assumptions on the memory contents that value relatedness may depend on.
However, value correctness may also depend on other assumptions.
Specifically, \stktokens{} has certain assumption on the seals used for return capabilities and closures.
For instance, a return seal must only be used to seal the return pointer of one specific return point.
Therefore, in addition to a relation on memory segments, some regions also carry a \emph{seal interpretation function} that relates the sealables that may be sealed with a given seal.
  \[
    \Seal \parfun \World \fun \Rel{\SealableCaps \times \SealableCaps}
  \]
%% Shared region has a seal indexed invariant that specifies what sealables can be sealed with a specific seal
  In \stktokens{}, once a seal has been used for a specific purpose (e.g.\ for sealing return capability pairs for a specific call site), it can never be reused for a different purpose.
  This is because there may still be copies of return capabilities out there, signed with the seal.
  This situation is similar to the situation for non-linear memory capabilities, so we only allow shared regions to carry seal interpretation functions, as we will see that those regions can never be revoked in future worlds.
\begin{multline*}
  \Regionh = 
  \{\pure \} \times (\World \fun \Rel{\MemSeg\times\MemSeg}) \times \\
  (\Seal \parfun \World \fun \Rel{\SealableCaps \times \SealableCaps})
\end{multline*}
We also refer to the seal interpretation function as the seal invariant, and we will refer to the memory relation as the memory invariant or just invariant when it is unambiguous.

@@

\subsubsection{Future worlds and revocation}
\label{subsubsec:ft-and-revocation}
% Memory evolves over time, we need to model this.
Very often, relatedness of two capabilities does not change if extra assumptions in the system are added.
For example, two related capabilities remain related when an extra invariant is added on unrelated memory, or when a stack frame that it does not reference is dropped.
In Kripke logical relations, such changes that do not invalidate the relatedness of values, are modelled by the future world relation $\future$.
The future world relation can also be thought of as the model of allowed changes in memory over time.
We will say that $W'$ is a future world of $W$ if $W' \future W$.

Relatedness of capabilities should be defined, so that it is monotone with respect to the future world relation.
In other words, relatedness of capabilities should only be predicated on assumptions in the world that are guaranteed to remain valid in future worlds, and the same holds for memory invariants.
As such, we require the world-indexed memory invariants to be monotone in the world:
\[
  \Regions = \left\{
  \begin{array}{l}
    \{\spatial \} \times (\World \monfun \Rel{\MemSeg \times \MemSeg}) \cup \\
    \{\spatialo \} \times (\World \monfun \Rel{\MemSeg \times \MemSeg}) \cup \\
    \{\revoked \}
  \end{array} \right.
\]
We make a similar change to $\Regionh$. The seal invariants must be monotone as well.

Kripke future world relations usually allow extending worlds with extra assumptions, or take steps in protocols that the system was designed to follow.
However, in our setting, we sometimes allow dropping assumptions, namely when linearity tells us that no value in the system depends on this assumption any more.
Specifically, if we have the only linear capability for a piece of memory, then we can be sure that there are no other capabilities for the same memory which makes it safe to repurpose the memory and drop or replace the previous assumption.
We mark dropped regions as revoked in the world following \citet{ahmed_2004} and \citet{thamsborg_2011} which for all intents and purposes corresponds to actually dropping the region.

We add a $\revoked$ tag to the spatial regions $\Regions$:
\[
  \Regions = \left\{
  \begin{array}{l}
    \{\spatial \} \times (\World \fun \Rel{\MemSeg \times \MemSeg}) \cup \\
    \{\spatialo \} \times (\World \fun \Rel{\MemSeg \times \MemSeg}) \cup \\
    \{\revoked \}
  \end{array} \right.
\]
The $\spatialo$-region may be depended on by a linear capability \emph{may} so we cannot allow it to be revoked.
On the other hand, no capability can depend on a $\spatial$-region, so it can be safely revoked.
% This may be confusing as a $\spatial$-region does not mean that you ``have'' the capability, but we just explained that having a linear capability means that you can repurpose the memory it points to.
% As we will see later, the logical relation splits worlds with respect to their ownership, so each capability can get its own world with unique ownership to depend on.
% A repurposed linear capability gets an entirely new world to depend on.
% The new world constructed for the repurposed linear capability would have a revoked region in place of the $\spatialo$-region, but the new world would not be a future world of the world the linear capability depended on before.
% After all, a future world has to respect old invariants, but a repurposed capability needs to depend on new invariants.
% While the repurposed capability gets a new world to depend on, all other capabilities in the system depend on the same invariants, so they must be valid with respect to worlds that respect the old invariants, i.e. future worlds.
% Further, the future worlds must be compatible (agree on everything but ownership) with the new world.
% This means that all the $\spatial$-regions must be replaced with $\revoked$ regions which is why $\spatial$ regions can be turned into $\revoked$ regions.

We define the future world relation in terms of a future region relation which is displayed in Figure~\ref{fig:ft-reg-rel}.
Apart from being revoked, a region can stay the same, or a $\spatial$ region can become $\spatialo$.
The latter allows us to reassign the $\spatialo$ region to some other world, when the linear capability owning the region is erased (our linear capabilities are actually affine).
% describe why "Spatial -> spatialo" ~ affine or at least why we have it?
\begin{figure}[htb]
  \centering
  \begin{mathpar}
  \inferrule{ r \in \Regions \cup \Regionh }{ r \future r}
  \and
  \inferrule{ }{ \revoked \future (\spatial,\_)}
  \and
  \inferrule{ }{ (\spatialo,H) \future (\spatial,H)}
\end{mathpar}
  \caption{Future region relation.}
  \label{fig:ft-reg-rel}
\end{figure}
The above repurposing is exemplified in \figurename~\ref{fig:world-example}.
The Figure displays two capabilities $c_{\var{normal}}$ and $c_{\var{linear}}$ that are normal and linear, respectively.
The linear capability is valid with respect to $W_1$ which has the necessary $\spatialo$ region, and the normal capability is valid with respect to $W_2$ which has a $\pure$ region.
The two worlds are compatible as $W_2$ has a $\spatial$ region that matches the $\spatialo$ region of $W_1$.
When the linear capability is repurposed, it must be reflected by the worlds $W_1'$ and $W_2'$.
In both new worlds, the $r_2$ region is replaced with a $\revoked$ region.
$W_1'$ has a new $\spatialo$ region at $r_3$, and $W_2'$ gets a matching $\spatial$ region.
The new world for the linear capability, $W_1'$, is not a future world of $W_1$ as it replaces a $\spatialo$ region with a $\revoked$ region which is not allowed by the future region relation.
The new world for the normal capability, $W_2'$, is a future world of $W_2$ as the future region relation allows $\spatial$ regions to be revoked.
The normal capability $c_{\var{normal}}$ remains valid in $W_2'$ as the $\pure$ region it depends on is monotone with respect to the future world relation.
\begin{figure}\centering

@@

\subsubsection{Joining worlds}
\label{subsubsec:joining-worlds}
The world serves multiple purposes as it is both a specification of memory contents as well as a specification of authority.
This is best seen in the operators used to join worlds.
First when we see the world as a memory specification, we have a pretty standard join $\uplus$ that simply requires the worlds to have different region names.\begin{definition}[World disjoint union $\uplus$]
  Given worlds $W_1$, $W_2$, $W$
  \[
    W_1 \uplus W_2 = W
    \text{ iff }
    \begin{array}[t]{l}
      \dom(\pwheap) = \dom(\pwheap[W_1]) \uplus \dom(\pwheap[W_2]) \tand \\
      \dom(\pwfree) = \dom(\pwfree[W_1]) \uplus \dom(\pwfree[W_2]) \tand \\
      \dom(\pwpriv) = \dom(\pwpriv[W_1]) \uplus \dom(\pwpriv[W_2]) \\
    \end{array}
  \]
\end{definition}
The $\uplus$ world join does not guarantee that the result is a sensible world with respect to authority or memory specification: it may contain regions with conflicting requirements for memory.
In \sectionname~\ref{subsubsec:mem-sat}, we define memory satisfaction which also acts as a well-formedness judgement.

When we view the world as a specification of authority, then the world join need to respect the region ownership.
That is, when we join the authority of two worlds, then the ownership of the two worlds should not overlap.
This is expressed by the $\oplus$ operator.
\begin{definition}[$\oplus$, disjoint union of ownership]
  $W_1 \oplus W_2 = W$ iff
  \[
  \begin{array}[t]{l}
    \dom(\pwheap) = \dom(\pwheap[W_1]) = \dom(\pwheap[W_2]) \tand \\
    \dom(\pwfree) = \dom(\pwfree[W_1]) = \dom(\pwfree[W_2]) \tand \\
    \dom(\pwpriv) = \dom(\pwpriv[W_1]) = \dom(\pwpriv[W_2]) \tand \\
    \forall r \in \dom(\pwheap) \ldotp \pwheap(r) = \pwheap[W_1](r) \oplus \pwheap[W_2](r) \tand \\
    \forall r \in \dom(\pwfree) \ldotp \pwfree(r) = \pwfree[W_1](r) \oplus \pwfree[W_2](r) \tand \\
    \forall r \in \dom(\pwpriv) \ldotp \pi_1(\pwpriv(r)) = \pi_1(\pwpriv[W_1](r)) \oplus \pi_1(\pwpriv[W_2](r))
  \end{array}
  \]
  where $\oplus$ for regions is defined as
\begin{align*}
  (\pure,H,H_{\mathrm{seal}}) \oplus (\pure,H,H_{\mathrm{seal}}) =  & \; (\pure,H,H_{\mathrm{seal}}) \\
  (\spatial,H) \oplus (\spatial,H) =  & \; (\spatial,H) \\
  \revoked \oplus \revoked = & \; \revoked \\
  (\spatialo,H) \oplus (\spatial,H) = & \; (\spatial,H) \oplus (\spatialo,H)\\
                                           =  & \; (\spatialo,H)
\end{align*}
\end{definition}
Like the $\uplus$ operator, the $\oplus$ operator does not guarantee that the resulting world is sensible.
